Generating Correlated Data for Omics Simulation

### 2025-05-09 manuscript/paper.qmd
Start rewrite of terminology

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing this $\Sigma$ presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
However, using $\widehat \Sigma$ is a poor choice for $\Sigma$.
Indeed, we consider a choice of $\Sigma$ to be successful if data sets of $n$ samples simulated using $\Sigma$ have sample covaraince matrices that are similar to $\widehat \Sigma$.
Counter-intuitively, data simulated using $\Sigma := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.

```{r}
#| label: fig-intro
#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma$ when estimated from the reference data set and when estimated from simulated data using $\\Sigma$ equal to $\\widehat \\Sigma$ of the reference data set ('simulated'). Eigenvalues are numbered by size. 100 genes and 12 samples were used for both reference and simulated data sets. Both reference and simulated data follow a multivariate normal distribution."
#| fig-width: 5
#| fig-height: 4

@@

One reason this occurs is that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
This limitation means that any data generated from the multivariate normal distribution of that $\Sigma$ will lie on a $n-1$-dimensional hyperplane.
Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$

We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma$, all which take the form
$$ \Sigma = D^2 + PP^T $$
where $D$ is a diagonal matrix capturing the injected independence and $PP^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
An additional advantage of this special form is that data may be generated very efficiently, see Methods, compared to arbitrary $\Sigma$ which typically require a Cholesky or eigenvalue decomposition, see Supplemental Figure S3.

@@

Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D^2 + P P^T$.

@@

The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ signular values.
This method computes $\Sigma$ such that $u_i^T \Sigma u_i = \lambda_i^2$, i.e. that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
Instead, we use the following:

@@

Steps 1 and 2 give that $u_i \Sigma u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
Step 4 ensures that $e_j \Sigma e_j^T = 1$ for $j = 1, \ldots, p$.

@@

The second method also makes use of PCA but has a different objective.
If $n$ samples are drawn from $N(0, \Sigma)$ then we want the variances of their PCA components to match those of the reference dataset.
Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma)$.
Then we want to choose $\Sigma$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.

Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma)$.
The spiked Wishart is the special case where $\Sigma$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
While our case has $\Sigma$ non-diagonal, $\Sigma$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
Therefore, the distribution of singular values is not affected by the assumption that $\Sigma$ is diagonal.
Moreover, for the form $\Sigma = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
This is a spiked Wishart distribution.

However, the spiked Wishart distribution also has no known analytic solution for the distribution of its eigenvalues either.
Therefore, we use an efficient sampling and stochastic gradient descent method that we recently described [@wishart].
Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma e_i^T \approx 1$.

@@

1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
3. Set $D = cI$

@@

The `corpcor` package [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] computes a James-Stein type shrinkage estimator for the covariance matrix.
For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D^2 + UWU^T$.

>>>>

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing the covariance matrix $\Sigma_{sim}$ to use during simulation presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma$ of the reference data set (possibly after some transformation of the data).
However, using $\widehat \Sigma$ is a poor choice for $\Sigma_{sim}$.
Indeed, we consider a choice of $\Sigma_{sim}$ to be successful if data sets of $n$ samples simulated using $\Sigma_{sim}$ have sample covaraince matrices $\widehat \Sigma_{sim}$ that are similar to $\widehat \Sigma$.
Counter-intuitively, data simulated using $\Sigma_{sim} := \widehat \Sigma$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma$, see @fig-intro.

```{r}
#| label: fig-intro
#| fig-cap: "Eigenvalues of the sample covariance matrix $\\widehat \\Sigma$ of reference data compared to $\\widehat \\Sigma_{sim}$ from from simulated data using $\\Sigma_{sim} = \\widehat \\Sigma$. 100 genes and 12 samples were used for both reference and the 20 simulated data sets. Both reference and simulated data follow a multivariate normal distribution. Eigenvalues are numbered from largest to smallest."
#| fig-width: 5
#| fig-height: 4

@@

One reason this occurs is that $\widehat \Sigma$ is low rank, specifically at most rank $n - 1$, which is less than $p$ in typical omics studies.
This means that any data generated using $\Sigma_{sim} = \widehat \Sigma$ will lie on a $n-1$-dimensional hyperplane.
Therefore, newly simulated samples will actually be linear combination of the samples of the reference data set.
Intuitively, this means that the entire probability distribution of possible $p$ dimensional data have been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference data set.
In contrast, we know that omics contains at least some level of random noise that is approximately independent from one gene to the next.
For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma$

We present three methods - called here PCA, spiked Wishart, and corpcor - for making this choice of $\Sigma_{sim}$, all which take the form
$$ \Sigma_{sim} = D^2 + UWU^T $$
where $D$ is a diagonal matrix capturing the injected independence and $UWU^T$ is a low-rank correlation matrix derived from $\widehat \Sigma$.
An additional advantage of this special form is that data may be generated very efficiently, Supplemental Figure S3 and see Methods, compared to arbitrary $\Sigma_{sim}$ which typically require a Cholesky or eigenvalue decomposition.

@@

Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{sim} := D^2 + P P^T$.

@@

The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ signular values.
This method computes $\Sigma_{sim}$ such that $u_i^T \Sigma_{sim} u_i = \lambda_i^2$, i.e. that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
Instead, we use the following:

@@

Steps 1 and 2 give that $u_i \Sigma_{sim} u_i^T = \lambda_i$ for $i = 1, \ldots, k$.
Step 4 ensures that $e_j \Sigma_{sim} e_j^T = 1$ for $j = 1, \ldots, p$.

@@

The second method also makes use of PCA but has a different objective.
If $n$ samples are drawn from $N(0, \Sigma_{sim})$ then we want the variances of their PCA components to match those of the reference dataset.
Specifically, let $\lambda_1, \ldots, \lambda_{n-1}$ be the ${n-1}$ non-zero singular values of $Z$ ($\lambda_n$ is always approximately zero due to the normalization procedure), and let $\lambda'_1, \ldots, \lambda'_{n-1}$ be the singular values of $Z'$ where $Z'$ has $n-1$ columns each iid $N(0, \Sigma_{sim})$.
Then we want to choose $\Sigma_{sim}$ such that $E[\lambda'_i] = \lambda_i$ for each $i$, where $E[Y]$ denotes the expectation of the random variable $Y$.

Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma_{sim})$.
The spiked Wishart is the special case where $\Sigma_{sim}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
Moreover, for the form $\Sigma_{sim} = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
This is a spiked Wishart distribution.

However, the spiked Wishart distribution also has no known analytic solution for the distribution of its eigenvalues either.
Therefore, we use an efficient sampling and stochastic gradient descent method that we recently described [@wishart].
Since the dataset has been normalized, $c$ will be close to one and $e_i \Sigma_{sim} e_i^T \approx 1$.

@@

1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma_{sim}$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
3. Set $D = cI$

@@

The `corpcor` package [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] computes a James-Stein type shrinkage estimator for the covariance matrix.
For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma_{sim} = D^2 + UWU^T$.
